---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
keep_md: true
---


## Question 1: Loading and preprocessing the data ##
We load the activity data from csv into dataframe activity0. The original data contains 17568 observations which is partially missing in variable steps. 

```{r}
activity0 <- read.csv("activity.csv")
str(activity0) # 17568 observations
```

We remove the rows with missing data and obtain data frame activity.
```{r}
wd  <- complete.cases(activity0)
activity = activity0[wd,] # 15264 left
str(activity)
```


##Question 2: What is mean total number of steps taken per day? ##
We use library dplyr to calculate the total number of steps per day. 

```{r}
library(dplyr)
resQ2 <- activity %>%
  group_by(date) %>%
  summarize(totalsteps = sum(steps,na.rm=TRUE))
```

The mean and median are 
```{r}
mean(resQ2$totalsteps)
median(resQ2$totalsteps)
```

Use library ggplot2 to create a histogram
```{r}
library(ggplot2)
qplot(totalsteps,data = resQ2,bins = 10,xlab = "Total number of steps",main = "Total number of steps per day [remove rows missing values]")
```

## Question 3: What is the average daily activity pattern? ##
Calculate the average number of steps taken averaged across all days in the dataset. 
```{r}
resQ3 <- activity %>%
  group_by(interval) %>%
  summarize(avsteps = mean(steps,na.rm=TRUE))
```

We find the maximum number of steps is on average made on interval 8:35
```{r}
maxsteps <- which.max(resQ3$avsteps)
resQ3$interval[maxsteps]
```

This brings the following lineplot 
```{r}
qplot(interval,avsteps, data = resQ3, ylab= "Average number of steps", main = "Average number of steps per 5 minute interval") + geom_line()
```

## Question 4: Imputing missing values ##
There were missing data in the original dataset. 

```{r}
wd  <- complete.cases(activity0)
sum(!wd) # 2304 rows with NA
```

Above we took out the rows with missing values. Here we impute values. We recognize that we cannot add mean/median from that specific day
as there are days which are completely NA. Thus we replace missing values with the mean for the 5 minute interval. 
Create dataframe activity_ext which picks the original number of steps when it is available, and  otherwise the mean value.

```{r}
activity_ext <- merge(activity0,resQ3,by.x="interval",by.y="interval")
activity_ext$steps2 = ifelse (is.na(activity_ext$steps), activity_ext$avsteps,activity_ext$steps)
```

And derive the total number of steps per day
```{r}
resQ4 <- activity_ext %>%
  group_by(date) %>%
  summarize(totalsteps = sum(steps2,na.rm=TRUE))
```


The mean and median have hardly changed from the above situation where we removed the rows
```{r}
mean(resQ4$totalsteps)
median(resQ4$totalsteps)
```

The histogram becomes
```{r}
qplot(totalsteps,data = resQ4,bins = 10,xlab = "Total number of steps",main = "Total number of steps per day [imputed values]")
```

Thus we conclude the impact of imputing missing data versus deleting the rows is limited.



## Q5 Are there differences in activity patterns between weekdays and weekends? ##
Add a new factor variable to our dataframe indicating whether it was weekday or weekend
```{r}
library(lubridate)
activity_ext$date = ymd(activity_ext$date) # make a date object

activity_ext$weekday = wday(activity_ext$date)
indWeekend = (activity_ext$weekday== 7 |  activity_ext$weekday== 1)
activity_ext$week_weekend[indWeekend] = "weekend"
activity_ext$week_weekend[!indWeekend] = "week"
activity_ext$week_weekend =as.factor(activity_ext$week_weekend)
```

Create panel plot with the average number of steps
```{r}
resQ5 <- activity_ext %>%
  group_by(interval,week_weekend) %>%
  summarize(avsteps = mean(steps,na.rm=TRUE))

qplot(interval,avsteps, data = resQ5,facets = .~week_weekend, ylab= "Average number of steps") + geom_line()
```
